---
title: "Global Attitudes Workshop"
subtitle: "Peer Group Analysis with Pew Spring 2024 Data"
author: "Bence Ságvári"
date: today
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 2
    number-sections: false
    embed-resources: true
    theme: cosmo
    code-fold: false
    df-print: paged
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true
#| echo: false
# Hidden initialisation chunk — students can ignore this.
```

---

# Before You Start {.unnumbered}

::: {.callout-important}
## 👥 How this session works — read before opening RStudio

This is a **90-minute group workshop**. You work in groups of 2–3. Every group has:

- **One Navigator** — the person with more R experience. Your job is to *explain*, not to type. Read the code out loud, explain what each line does, and let your partner(s) drive. Resist the urge to grab the keyboard.
- **One or two Drivers** — you type the code. If you don't understand a line, ask your Navigator before running it. Your goal is to understand every line, not just copy it.

**Switch roles** at the halfway point (roughly after Exercise 3).

There are **6 exercises** designed for 90 minutes. Exercises 1–3 are more guided; 4–6 give you more freedom. Don't rush — a thoughtful answer to 4 exercises is worth more than a hurried pass through all 6.

**The output of each exercise is:**

1. A working plot or table that renders in the document
2. 2–3 written sentences of interpretation in the **💬 Your group's interpretation** box
3. A brief discussion — use the discussion questions as a prompt

You do **not** need to submit a rendered HTML for this session. Work together, discuss, and be ready to share one finding with the class at the end.

**Tip for running code:** Use the green ▶ button or **Ctrl+Shift+Enter** to run a single chunk.  
Use the **⬇ Run All Chunks Above** button whenever you see an error — it reloads the whole environment.
:::

---

# Session Setup {#sec-setup}

::: {.callout-note collapse="true"}
## 📦 What these packages do — Navigator, explain this to your Driver

| Package | What it does in this notebook |
|---------|------------------------------|
| `tidyverse` | Filter rows, create new columns, make plots with ggplot2 |
| `haven` | Read the Pew `.sav` (SPSS) file — it stores value labels alongside numbers |
| `srvyr` | Apply survey weights correctly — without this, our percentages are wrong |
| `knitr` | Format tables neatly when the document renders |
| `scales` | Format axis labels as percentages |

**Navigator talking point:** Ask your Driver — *why can't we just use `mean()` on survey data? What goes wrong?*
:::

```{r}
#| label: load-packages
#| eval: true

# ── Install any packages that are not yet on this machine ────────────────────
# installed.packages() returns a matrix of everything currently installed.
# We check which of our required packages are missing, then install only those.
# This avoids reinstalling packages that are already present.
packages <- c("tidyverse", "haven", "srvyr", "knitr", "scales")
missing  <- packages[!packages %in% installed.packages()[, "Package"]]
if (length(missing) > 0) install.packages(missing)

# ── Load all packages into the current session ───────────────────────────────
# lapply() applies library() to every element of the vector.
# invisible() suppresses the list of return values (keeps the console clean).
# character.only = TRUE is needed because library() normally expects an unquoted name.
invisible(lapply(packages, library, character.only = TRUE))

# ── Set a default ggplot2 theme for all plots in this document ────────────────
# theme_set() applies globally — we don't need to repeat theme_minimal() in every plot.
# base_size = 12 sets the default font size for axis labels and titles.
theme_set(theme_minimal(base_size = 12))
```

```{r}
#| label: load-data
#| eval: true
#| cache: true

# ── Download the Pew Spring 2024 Global Attitudes dataset ─────────────────────
# The data is hosted on the course SharePoint. We download it to a temporary
# file so it doesn't clutter your project folder.
# tempfile() creates a path to a temporary location; fileext = ".sav" ensures
# the file is recognised as SPSS format.
data_link <- "https://tkmtahu-my.sharepoint.com/:u:/g/personal/sagvari_bence_tk_hu/IQCc8k4_cII9TIlJmZQULNd6ATVy7R_EEkU5AwRBspfjP_A?download=1"
tmp <- tempfile(fileext = ".sav")

# mode = "wb" means "write binary" — required for non-text files like .sav
download.file(data_link, tmp, mode = "wb", quiet = TRUE)

# haven::read_spss() reads SPSS .sav files.
# It preserves both the numeric codes AND the human-readable labels.
pew <- haven::read_spss(tmp)

# Quick sanity check — print the dimensions
cat("Dataset loaded:", nrow(pew), "respondents,", ncol(pew), "variables\n")
```

```{r}
#| label: clean-data
#| eval: true

# ── Country codes for Europe + Canada ────────────────────────────────────────
# In the Pew dataset, 'country' is stored as a numeric code.
# We list the codes we want to keep (see data dictionary for the full list).
eu_can_codes <- c(2,   # Canada
                  3,   # France
                  4,   # Germany
                  5,   # Greece
                  6,   # Hungary
                  7,   # Italy
                  8,   # Netherlands
                  9,   # Poland
                  10,  # Spain
                  11,  # Sweden
                  12)  # United Kingdom

# ── Build the cleaned analysis dataset ───────────────────────────────────────
pew_clean <- pew |>

  # Keep only European + Canadian respondents
  filter(country %in% eu_can_codes) |>

  # zap_missing() converts SPSS "system missing" and user-defined missing codes
  # (e.g. 8 = "Don't know", 9 = "Refused") into proper R NA values.
  # Without this step, respondents who refused would appear to have answered "9"
  # on a 4-point scale, completely distorting any calculation.
  zap_missing() |>

  mutate(

    # ── Convert numeric codes to readable factor labels ──────────────────────
    # as_factor() uses the value labels stored in the SPSS file.
    # Result: country becomes "Canada", "France", etc. instead of 2, 3, etc.
    country_f = as_factor(country),
    gender_f  = as_factor(gender),

    # ── Reverse-code the attitude scales ─────────────────────────────────────
    # Most Pew scales run 1 = most positive … 4 = most negative.
    # We reverse them so that HIGHER values always mean MORE of the concept.
    # Formula: reversed = (max_value + 1) - original
    # Example: dem_sat_r = 5 - satisfied_democracy
    #   Original 1 (Very satisfied)     → reversed 4
    #   Original 4 (Not at all satisfied) → reversed 1

    # Democracy satisfaction (original: 1=Very satisfied … 4=Not at all)
    dem_sat_r   = 5 - satisfied_democracy,  # reversed: 4 = most satisfied

    # National economic situation (original: 1=Very good … 4=Very bad)
    econ_r      = 5 - econ_sit,             # reversed: 4 = economy very good

    # Importance of religion (original: 1=Very important … 4=Not at all)
    rel_imp_r   = 5 - religion_import,      # reversed: 4 = very important

    # Press freedom importance (original: 1=Very important … 4=Not at all)
    press_imp_r = 5 - press_freedom,        # reversed: 4 = very important

    # Support for autocracy (original: 1=Very good … 4=Very bad way to govern)
    autocracy_r = 5 - polsys_autocracy,     # reversed: 4 = strong support

    # Support for representative democracy (original: 1=Very good … 4=Very bad)
    republic_r  = 5 - polsys_republic,      # reversed: 4 = strong support

    # ── Create binary (0/1) indicator variables ───────────────────────────────
    # A binary variable's mean = the proportion of 1s.
    # as.integer() converts TRUE/FALSE to 1/0.

    # 1 if respondent is satisfied or very satisfied with democracy (codes 1 or 2)
    satisfied   = as.integer(satisfied_democracy <= 2),

    # 1 if respondent thinks children will be financially WORSE OFF than parents
    pessimist   = as.integer(children_betteroff == 2),

    # 1 if respondent would be comfortable if their child came out as gay/lesbian
    gay_comfort = as.integer(child_gay == 1),

    # 1 if religion is "Very important" in the respondent's life (code 1 only)
    rel_high    = as.integer(religion_import == 1),

    # Political self-placement on a 0 (extreme left) to 6 (extreme right) scale
    # We just rename it for clarity; no transformation needed
    ideology    = political_scale2
  )

# ── Create the survey design object ─────────────────────────────────────────
# as_survey_design() wraps pew_clean in a special structure that "remembers"
# which column contains the survey weights.
# After this step, every estimate computed with survey_mean() or survey_total()
# will automatically apply the weights — no need to pass w = weight every time.
# ids = 1 means "no cluster variable" (Pew does not release one for public use).
pew_svy <- pew_clean |>
  as_survey_design(ids = 1, weights = weight)

cat("Clean dataset ready:", nrow(pew_clean), "respondents across",
    n_distinct(pew_clean$country_f), "countries\n")
```

::: {.callout-tip collapse="true"}
## 🧭 Navigator: quick orientation before Exercise 1

Run `glimpse(pew_clean)` in the console. Point out to your Driver:

- The `_r` columns are reversed scales — higher always means "more" of the concept
- The binary columns (`satisfied`, `pessimist`, `gay_comfort`) are 0/1 — the mean of a 0/1 variable is the proportion
- `pew_svy` is the survey design object — use this in all `survey_mean()` calls, not `pew_clean`
:::

---

# Exercise 1 — Does Weighting Actually Matter? {#ex1}

::: {.callout-note}
## 🗺 Mission · ⏱ ~12 min

Surveys are not random samples of the population — some groups respond more readily than others. The `weight` variable corrects for this. But how much does it actually change our estimates?

**Your task:** Compare the *unweighted* and *weighted* mean democracy satisfaction score for each country. Find out which country's estimate changes the most when weights are applied.
:::

::: {.callout-warning collapse="true"}
## 💡 Hint — Navigator reads this aloud, Driver types

`mean()` ignores weights and treats every respondent equally.  
`weighted.mean(x, w = weight)` gives each respondent a weight proportional to how many people in the real population they represent.  
To compare both: compute them in one `summarise()` call using `group_by(country_f)`.
:::

```{r}
#| label: ex1-compare-weights

# ── Compare unweighted vs. weighted means by country ─────────────────────────
# We use pew_clean (not pew_svy) here so we can call both mean() and
# weighted.mean() side by side in the same summarise() call.

weight_check <- pew_clean |>

  # Drop rows where satisfaction or weight is missing
  filter(!is.na(dem_sat_r), !is.na(weight)) |>

  # Compute one row per country
  group_by(country_f) |>

  summarise(

    # Unweighted mean: every respondent counts equally, regardless of weight
    unweighted = mean(dem_sat_r, na.rm = TRUE),

    # Weighted mean: each respondent is scaled by their survey weight
    # w = weight tells R which column contains the weight values
    weighted   = weighted.mean(dem_sat_r, w = weight, na.rm = TRUE),

    # The difference between the two — positive means weighting pushed estimate up
    diff       = weighted - unweighted,

    # n() counts the number of rows (respondents) in each group
    n          = n(),

    .groups    = "drop"  # remove the grouping after summarise (good practice)

  ) |>

  # Sort by absolute difference — countries where weighting matters most appear first
  arrange(desc(abs(diff)))

# kable() formats the data frame as a clean HTML table
kable(weight_check,
      digits    = 3,
      col.names = c("Country", "Unweighted mean", "Weighted mean", "Difference", "N"),
      caption   = "Effect of weighting on democracy satisfaction (scale: 1=Not at all satisfied, 4=Very satisfied)")
```

```{r}
#| label: ex1-plot

# ── Scatter plot: weighted vs. unweighted mean per country ────────────────────
# Points ON the dashed diagonal line = weighting made no difference.
# Points BELOW the line = weighting pushed the estimate down.
# Points ABOVE the line = weighting pushed the estimate up.

ggplot(weight_check,
       aes(x = unweighted,    # unweighted mean on the horizontal axis
           y = weighted,      # weighted mean on the vertical axis
           label = country_f  # country name used by geom_text below
       )) +

  # Draw a y = x reference line — any point on this line has diff = 0
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey60") +

  # One dot per country; size encodes the magnitude of the difference
  geom_point(aes(size = abs(diff)),
             colour = "#1B3A5C", alpha = 0.8) +

  # Country labels, shifted slightly upward so they don't overlap the dots
  geom_text(nudge_y = 0.012, size = 3.5, colour = "grey30") +

  # scale_size_area() maps the size aesthetic to dot area (not radius),
  # which is more perceptually accurate. guide = "none" hides the size legend.
  scale_size_area(max_size = 9, guide = "none") +

  labs(
    title    = "Weighted vs. unweighted democracy satisfaction",
    subtitle = "Points off the dashed line = weighting changes the estimate",
    x        = "Unweighted mean",
    y        = "Weighted mean",
    caption  = "Source: Pew Research Center Global Attitudes Spring 2024\nScale: 1 = Not at all satisfied, 4 = Very satisfied"
  )
```

::: {.callout-caution}
## 💬 Your group's interpretation — write 2–3 sentences here

*Which country's estimate changes the most? Are the differences large enough to matter substantively? What might explain why some countries are further from the diagonal than others?*

**Group answer:** *[write here]*
:::

::: {.callout-tip collapse="true"}
## 🚀 Go further — for groups that finish early

Repeat the comparison but for **economic pessimism** (`pessimist`). Does weighting matter more or less for this variable? Can you explain why?
:::

---

# Exercise 2 — Who Trusts Their Democracy? {#ex2}

::: {.callout-note}
## 🗺 Mission · ⏱ ~15 min

Democracy satisfaction varies enormously across Europe. Sweden and the Netherlands are near the top; Greece and France near the bottom. But are these differences explained by *where* people live, or by *who* they are — their gender, for instance?

**Your task:**  
1. Build a ranked bar chart of weighted democracy satisfaction by country, with 95% confidence interval error bars.  
2. Then break it down by gender — does the gender gap look the same in every country?
:::

::: {.callout-warning collapse="true"}
## 💡 Hint — Navigator reads this aloud

`survey_mean(x, na.rm = TRUE)` returns both the weighted proportion AND its standard error in one call.  
When you assign it like `pct_satisfied = survey_mean(...)`, srvyr automatically creates two columns: `pct_satisfied` (the estimate) and `pct_satisfied_se` (the standard error).  
A 95% confidence interval is: estimate ± 1.96 × SE.

For the gender breakdown: add `gender_f` to the `group_by()` call. Use `facet_wrap(~gender_f)` or `aes(fill = gender_f)` with `position = "dodge"`.
:::

```{r}
#| label: ex2-country-ranking

# ── Weighted % satisfied by country, with standard errors ────────────────────
# We use pew_svy (the survey design object) so that survey_mean() applies
# weights and computes design-correct standard errors automatically.

dem_by_country <- pew_svy |>

  # Drop rows without a valid satisfaction response
  filter(!is.na(satisfied)) |>

  # One group per country
  group_by(country_f) |>

  summarise(
    # survey_mean() returns the weighted proportion.
    # Multiplying by 100 converts to percentage — srvyr multiplies BOTH
    # the estimate and the SE column by 100 in a single step.
    # Result: two new columns named `pct_satisfied` and `pct_satisfied_se`
    pct_satisfied = survey_mean(satisfied, na.rm = TRUE, vartype = "se") * 100,
    .groups       = "drop"
  ) |>

  # Sort highest satisfaction first (top of the bar chart = Sweden)
  arrange(desc(pct_satisfied))

# Print as a table with three columns: Country | % Satisfied | SE
kable(dem_by_country,
      digits    = 1,
      col.names = c("Country", "% Satisfied", "SE (pp)"),
      caption   = "% satisfied or very satisfied with democracy (weighted)")
```

```{r}
#| label: ex2-bar-chart

# ── Horizontal bar chart with 95% confidence interval error bars ──────────────

ggplot(dem_by_country,
       aes(
         # reorder() sorts countries by their satisfaction percentage
         # so the highest bar appears at the top after coord_flip()
         x = reorder(country_f, pct_satisfied),
         y = pct_satisfied
       )) +

  # Filled bars
  geom_col(fill = "#1B3A5C", width = 0.7) +

  # Error bars showing the 95% confidence interval.
  # The CI is computed as: estimate ± 1.96 × SE.
  # pct_satisfied_se is already in percentage points (×100 was applied above),
  # so we do NOT multiply it by 100 again here.
  geom_errorbar(
    aes(
      ymin = pct_satisfied - 1.96 * pct_satisfied_se,  # lower bound of 95% CI
      ymax = pct_satisfied + 1.96 * pct_satisfied_se   # upper bound of 95% CI
    ),
    width = 0.25, colour = "grey50"
  ) +

  # Percentage labels just to the right of each bar
  geom_text(aes(label = paste0(round(pct_satisfied, 0), "%")),
            hjust = -0.2, size = 3.5) +

  # Flip axes: countries on the y-axis, percentages on the x-axis
  coord_flip() +

  scale_y_continuous(limits = c(0, 100)) +

  labs(
    title    = "Satisfaction with how democracy works",
    subtitle = "% 'Very' or 'Somewhat' satisfied — error bars show 95% CI",
    x        = NULL, y = "Percentage",
    caption  = "Source: Pew Research Center Global Attitudes Spring 2024"
  )
```

```{r}
#| label: ex2-gender-gap

# ── ✏️ Your turn — break down satisfaction by gender ─────────────────────────
# Navigator: explain what needs to change in the group_by() call.
# Adding gender_f as a second grouping variable gives one row per
# country × gender combination.

# Driver: modify the code below
# 1. Add gender_f to group_by()
# 2. Filter out rows where gender_f is NA (use !is.na(gender_f))
# 3. Use facet_wrap(~gender_f) OR aes(fill = gender_f) with position = "dodge"

dem_by_gender <- pew_svy |>
  filter(!is.na(satisfied), !is.na(gender_f)) |>
  group_by(country_f, gender_f) |>      # group by BOTH country and gender
  summarise(
    pct_satisfied = survey_mean(satisfied, na.rm = TRUE, vartype = "se") * 100,
    .groups       = "drop"
  )

# ── YOUR PLOT CODE HERE ───────────────────────────────────────────────────────
# Suggested structure:
# ggplot(dem_by_gender, aes(x = reorder(country_f, pct_satisfied),
#                           y = pct_satisfied, fill = gender_f)) +
#   geom_col(position = "dodge", width = 0.7) +
#   coord_flip() + ...
```

::: {.callout-caution}
## 💬 Your group's interpretation — write 2–3 sentences here

*Which country has the highest democracy satisfaction? How large is the gender gap, and is it consistent across countries? Does anything in the results surprise you?*

**Group answer:** *[write here]*
:::

::: {.callout-tip collapse="true"}
## 🚀 Go further

Add error bars to the gender comparison. In which countries is the gender gap statistically meaningful — i.e. the 95% CI bands for men and women do not overlap?
:::

---

# Exercise 3 — The Economic Mood {#ex3}

::: {.callout-note}
## 🗺 Mission · ⏱ ~15 min

Europe is experiencing persistent economic anxiety. But pessimism varies enormously — Poland is relatively optimistic, France is deeply pessimistic. In this exercise you will visualise both the **current economic situation** and **intergenerational pessimism** side by side, and look for countries where the two tell different stories.

**Your task:**  
1. Build a grouped bar chart: one colour for % rating economy as good, another for % saying children will be worse off.  
2. Look for countries where the two patterns diverge — high economic satisfaction but also high pessimism, or vice versa.
:::

::: {.callout-warning collapse="true"}
## 💡 Hint — Navigator reads this aloud

Compute both variables in a single `summarise()` call, then use `pivot_longer()` to reshape the data from wide to long format — one row per country × indicator combination. This is the standard preparation step before `facet_wrap()` or grouped bars.
:::

```{r}
#| label: ex3-economic-mood

# ── Compute two economic indicators simultaneously ────────────────────────────
# We compute both indicators in one summarise() call to keep the code compact.
# Each survey_mean() call returns TWO columns: the estimate and the SE.
# With two calls we get four new columns total:
#   pct_econ_good, pct_econ_good_se, pct_pessimist, pct_pessimist_se

econ_mood <- pew_svy |>

  filter(!is.na(weight)) |>

  group_by(country_f) |>

  summarise(

    # % who rate the current economic situation as "Very good" or "Somewhat good"
    # econ_sit codes: 1 = Very good, 2 = Somewhat good → condition: <= 2
    pct_econ_good = survey_mean(econ_sit <= 2,           na.rm = TRUE, vartype = "se") * 100,

    # % who expect their children to be WORSE OFF financially than their parents
    # children_betteroff codes: 1 = Better off, 2 = Worse off, 3 = Same → condition: == 2
    pct_pessimist = survey_mean(children_betteroff == 2, na.rm = TRUE, vartype = "se") * 100,

    .groups = "drop"

  ) |>

  arrange(desc(pct_econ_good))  # sort by economic satisfaction (highest first)

# Print the full table (SE columns included)
kable(econ_mood,
      digits    = 1,
      col.names = c("Country",
                    "% Economy good", "SE",
                    "% Kids worse off", "SE"),
      caption   = "Economic attitudes by country (weighted %)")
```

```{r}
#| label: ex3-plot

# ── Reshape to long format for a grouped bar chart ────────────────────────────
# Currently econ_mood is in WIDE format: one row per country, two value columns.
# ggplot2 wants LONG format for grouped/faceted bars: one row per country × indicator.
# pivot_longer() does this transformation.

econ_long <- econ_mood |>

  # Keep only the point-estimate columns (drop the SE columns for the plot)
  select(country_f, pct_econ_good, pct_pessimist) |>

  pivot_longer(
    cols      = c(pct_econ_good, pct_pessimist), # the two columns to stack
    names_to  = "indicator",   # new column that stores which variable this row came from
    values_to = "pct"          # new column that stores the actual percentage value
  ) |>

  # Replace the raw column names with readable labels for the chart legend
  mutate(
    indicator = case_match(
      indicator,
      "pct_econ_good" ~ "% Rating economy as good",
      "pct_pessimist" ~ "% Expecting children to be worse off"
    )
  )

ggplot(econ_long,
       aes(
         # Sort countries so the most economically satisfied appears at the top
         x    = reorder(country_f, pct),
         y    = pct,
         fill = indicator   # colour encodes which indicator this bar represents
       )) +

  # position = "dodge" places bars for the same country side by side (not stacked)
  geom_col(position = "dodge", width = 0.7) +

  coord_flip() +  # flip axes so country names are readable on the y-axis

  # Custom colours: teal for economic optimism, red for pessimism
  scale_fill_manual(values = c("#1A7A7A", "#C0392B")) +

  scale_y_continuous(
    limits = c(0, 100),
    # label_percent(scale = 1) formats numbers as "50%" without multiplying by 100
    labels = scales::label_percent(scale = 1)
  ) +

  labs(
    title    = "Current economic mood vs. intergenerational pessimism",
    subtitle = "Weighted percentages — Pew Spring 2024",
    x        = NULL, y = NULL, fill = NULL,
    caption  = "Source: Pew Research Center Global Attitudes Spring 2024"
  ) +

  theme(legend.position = "top")
```

::: {.callout-caution}
## 💬 Your group's interpretation — write 2–3 sentences here

*Which country shows the biggest gap between current economic mood and intergenerational pessimism? What might explain this divergence? Is France's pessimism surprising given its economic situation?*

**Group answer:** *[write here]*
:::

---

> ### ⏸ Halfway pause — switch roles now
>
> If you haven't switched Driver and Navigator yet, do it now. The new Navigator should look back at Exercises 1–3 and explain to the new Driver: what does `survey_mean()` do, and why do we use `pew_svy` instead of `pew_clean`?

---

# Exercise 4 — Religion and Tolerance {#ex4}

::: {.callout-note}
## 🗺 Mission · ⏱ ~15 min

Are more religious countries also less socially tolerant? This is a classic question in the sociology of religion, and the Pew data lets us test it directly across 11 European countries and Canada.

**Your task:**  
1. For each country, compute: (a) % saying religion is very important; (b) % comfortable with having a gay child.  
2. Make a **scatter plot** with one dot per country and country name labels.  
3. Discuss: is the relationship linear? Are there outliers?
:::

::: {.callout-warning collapse="true"}
## 💡 Hint — Navigator reads this aloud

Use `group_by(country_f) |> summarise(...)` on `pew_svy`.  
For the scatter: `geom_point()` + `geom_text(aes(label = country_f), nudge_y = 1.5)`.  
Add a linear trend line with `geom_smooth(method = "lm", se = FALSE)`.  
The key interpretation question: does the trend hold for *all* countries, or is it driven by one or two outliers?
:::

```{r}
#| label: ex4-religion-tolerance

# ── Compute religiosity and tolerance indicators by country ───────────────────
# Both variables are binary (0/1), so survey_mean() gives the weighted proportion.
# rel_high = 1 if religion is "Very important" (religion_import == 1)
# gay_comfort = 1 if respondent would be comfortable if child came out as gay

rel_tol <- pew_svy |>

  filter(!is.na(weight)) |>

  group_by(country_f) |>

  summarise(

    # Weighted % for whom religion is "very important" in daily life
    pct_religious   = survey_mean(rel_high,    na.rm = TRUE, vartype = "se") * 100,

    # Weighted % who would be comfortable if a child came out as gay/lesbian
    pct_gay_comfort = survey_mean(gay_comfort, na.rm = TRUE, vartype = "se") * 100,

    .groups = "drop"

  )

# ── Scatter plot: one dot per country ────────────────────────────────────────
ggplot(rel_tol,
       aes(
         x     = pct_religious,    # religiosity on the horizontal axis
         y     = pct_gay_comfort,  # tolerance on the vertical axis
         label = country_f         # used by geom_text for country labels
       )) +

  # Draw a linear regression line (method = "lm") without confidence band (se = FALSE)
  # This shows the overall trend across countries
  geom_smooth(method = "lm", se = FALSE,
              colour = "grey70", linetype = "dashed") +

  # One dot per country, navy colour
  geom_point(size = 3, colour = "#1B3A5C") +

  # Country name labels, shifted slightly upward to avoid overlapping the dots
  geom_text(nudge_y = 2, size = 3.5, colour = "grey30") +

  # Format x-axis as percentages (scale = 1 means no ×100 conversion)
  scale_x_continuous(
    labels = scales::label_percent(scale = 1),
    limits = c(0, 40)
  ) +
  scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    limits = c(0, 25)
  ) +

  labs(
    title    = "Religion importance vs. comfort with LGBTQ+ family members",
    subtitle = "Each dot = one country — weighted estimates",
    x        = "% saying religion is 'very important' in life",
    y        = "% comfortable if a child came out as gay/lesbian",
    caption  = "Source: Pew Research Center Global Attitudes Spring 2024"
  )
```

::: {.callout-caution}
## 💬 Your group's interpretation — write 2–3 sentences here

*Describe the overall pattern. Are there any outliers — countries that don't fit the trend? What does this suggest about the relationship between religiosity and social tolerance in Europe?*

**Group answer:** *[write here]*
:::

::: {.callout-tip collapse="true"}
## 🚀 Go further

Add a third dimension: colour the points by `pct_pessimist` from Exercise 3. Does economic anxiety independently predict lower tolerance, above and beyond religiosity?

```r
# Step 1: join rel_tol with econ_mood (keep only point estimates, drop SE cols)
rel_tol_eco <- rel_tol |>
  left_join(
    econ_mood |> select(country_f, pct_pessimist),
    by = "country_f"
  )

# Step 2: add colour aesthetic
# ggplot(rel_tol_eco, aes(..., colour = pct_pessimist)) +
#   geom_point(size = 4) +
#   scale_colour_gradient(low = "#74add1", high = "#d73027",
#                         name = "% pessimist")
```
:::

---

# Exercise 5 — Who Wants a Strong Leader? {#ex5}

::: {.callout-note}
## 🗺 Mission · ⏱ ~15 min

Support for autocracy — a strong leader who governs without parliament or courts — is not uniformly distributed. It tends to be higher among older respondents, those on the right of the political spectrum, and those dissatisfied with democracy.

**Your task:**  
1. Compute support for autocracy and for representative democracy side by side, for the countries where data are available.  
2. Check whether political ideology (left–right scale) predicts support for autocracy. Split respondents into three ideology groups and compare.
:::

::: {.callout-warning collapse="true"}
## 💡 Hint — Navigator reads this aloud

`polsys_autocracy` was only asked in some countries — check coverage first.  
For the ideology breakdown: use `case_when()` to create three groups (Left / Centre / Right) from the 0–6 scale, then `group_by(country_f, ideo_group)`.  

Note: the condition for "support" uses the *original* variable: codes 1 ("Very good") or 2 ("Somewhat good") mean the respondent supports that system, so the condition is `<= 2`.
:::

```{r}
#| label: ex5-coverage-check

# ── Check which countries have data for the political systems questions ────────
# Before computing statistics, always verify how much data you actually have.
# sum(!is.na(x)) counts the number of non-missing values in column x.

pew_clean |>
  group_by(country_f) |>
  summarise(
    n_autocracy = sum(!is.na(polsys_autocracy)),  # rows with valid autocracy response
    n_republic  = sum(!is.na(polsys_republic)),   # rows with valid republic response
    .groups     = "drop"
  ) |>
  filter(n_autocracy > 0) |>  # keep only countries where the question was asked
  kable(caption = "Coverage of political systems questions by country")
```

```{r}
#| label: ex5-polsys-comparison

# ── Weighted support for autocracy vs. representative democracy ───────────────

polsys <- pew_svy |>

  # Keep rows where at least one of the two polsys questions was answered
  # The | (OR) operator keeps rows where EITHER condition is true
  filter(!is.na(polsys_autocracy) | !is.na(polsys_republic)) |>

  group_by(country_f) |>

  summarise(

    # % rating a strong leader (autocracy) as "Very good" or "Somewhat good"
    # Original codes: 1 = Very good, 2 = Somewhat good → condition: <= 2
    pct_autocracy = survey_mean(polsys_autocracy <= 2, na.rm = TRUE, vartype = "se") * 100,

    # % rating representative democracy as "Very good" or "Somewhat good"
    pct_republic  = survey_mean(polsys_republic  <= 2, na.rm = TRUE, vartype = "se") * 100,

    .groups = "drop"

  )

# ── Reshape to long format and plot ──────────────────────────────────────────
polsys |>

  # Drop the SE columns — we don't need them for this plot
  select(country_f, pct_autocracy, pct_republic) |>

  # Stack the two estimate columns into a single 'pct' column
  pivot_longer(
    cols      = c(pct_autocracy, pct_republic),
    names_to  = "system",
    values_to = "pct"
  ) |>

  # Replace raw column names with readable labels for the legend
  # case_match() is the modern replacement for the deprecated recode()
  mutate(
    system = case_match(
      system,
      "pct_autocracy" ~ "Strong leader (no parliament/courts)",
      "pct_republic"  ~ "Representative democracy"
    )
  ) |>

  ggplot(aes(x = country_f, y = pct, fill = system)) +

  # Grouped bars (side by side, not stacked)
  geom_col(position = "dodge", width = 0.6) +

  scale_fill_manual(values = c(
    "Strong leader (no parliament/courts)" = "#C0392B",
    "Representative democracy"             = "#1A7A7A"
  )) +

  scale_y_continuous(
    limits = c(0, 100),
    labels = scales::label_percent(scale = 1)
  ) +

  labs(
    title    = "Support for political systems",
    subtitle = "% rating each system as 'very good' or 'somewhat good' (weighted)",
    x        = NULL, y = NULL, fill = NULL,
    caption  = "Source: Pew Spring 2024 · Countries with available data only"
  ) +

  theme(legend.position = "top")
```

```{r}
#| label: ex5-ideology-breakdown

# ── ✏️ Your turn — does political ideology predict autocracy support? ─────────
# Navigator: explain case_when() before your Driver types.
# case_when() is like a chain of if-else conditions:
#   if ideology <= 2 → assign "Left (0–2)"
#   if ideology >= 4 → assign "Right (4–6)"
#   otherwise (TRUE) → assign "Centre (3)"

ideo_autocracy <- pew_svy |>

  # Keep only rows with valid autocracy and ideology values
  filter(!is.na(polsys_autocracy), !is.na(ideology)) |>

  # Create the ideology group variable BEFORE group_by
  # (mutate inside a survey design object is allowed)
  mutate(
    ideo_group = case_when(
      ideology <= 2 ~ "Left (0–2)",    # left-leaning respondents
      ideology >= 4 ~ "Right (4–6)",   # right-leaning respondents
      TRUE          ~ "Centre (3)"     # TRUE is the catch-all — everyone else
    )
  ) |>

  # Group by BOTH country and ideology group → one row per combination
  group_by(country_f, ideo_group) |>

  summarise(
    # % who consider autocracy "Very good" or "Somewhat good"
    pct_support = survey_mean(polsys_autocracy <= 2, na.rm = TRUE, vartype = "se") * 100,
    .groups     = "drop"
  )

# ── Faceted bar chart — one panel per country ─────────────────────────────────
ggplot(ideo_autocracy,
       aes(x = ideo_group,   # ideology group on the x-axis within each panel
           y = pct_support,  # % support on the y-axis
           fill = ideo_group # colour encodes the ideology group
       )) +

  geom_col(width = 0.6) +

  # facet_wrap() creates one separate mini-plot (panel) per country
  # ~country_f means "split by the country_f variable"
  facet_wrap(~country_f) +

  # Manual colour mapping: blue for left, grey for centre, red for right
  scale_fill_manual(
    values = c(
      "Left (0–2)"  = "#4393c3",
      "Centre (3)"  = "#d1d1d1",
      "Right (4–6)" = "#d73027"
    ),
    guide = "none"  # suppress the legend (colours are self-explanatory)
  ) +

  scale_y_continuous(
    limits = c(0, 60),
    labels = scales::label_percent(scale = 1)
  ) +

  labs(
    title    = "Support for autocracy by political ideology",
    subtitle = "% rating strong-leader rule as 'very' or 'somewhat good' (weighted)",
    x        = NULL, y = NULL,
    caption  = "Source: Pew Spring 2024"
  )
```

::: {.callout-caution}
## 💬 Your group's interpretation — write 2–3 sentences here

*Is support for autocracy higher on the right, the left, or in the centre? Does the pattern look the same in Hungary and Poland? What surprised your group?*

**Group answer:** *[write here]*
:::

---

# Exercise 6 — Your Group's Question {#ex6}

::: {.callout-note}
## 🗺 Mission · ⏱ ~18 min

This is the open exercise. Your group picks a research question, builds the analysis, and prepares one finding to share with the class.

**Step 1:** Agree on a question. Some starting points:

- Does **misinformation concern** vary by country — and is it higher among people who are less satisfied with democracy?
- Is support for **gender equality** (as an essential value) related to religiosity?
- Does **political ideology** predict economic pessimism, and does this relationship differ across countries?
- Pick two variables from the reference table below — formulate a hypothesis and test it.

**Step 2:** Build the analysis using the scaffold below.

**Step 3:** Write your interpretation and prepare to present one plot to the class in 2–3 minutes.
:::

::: {.callout-warning collapse="true"}
## 💡 Variable reference — useful variables for free exploration

| Variable (in `pew_clean`) | What it measures | Scale direction |
|--------------------------|-----------------|----------------|
| `prob_misinformation` | Perceived severity of misinformation | 1 = Very big problem … 4 = Not a problem |
| `value_gender` | Gender equality as a national value | 1 = Essential … 3 = Not important |
| `trust_gov` | Trust in national government | 1 = A lot … 4 = Not at all |
| `conflict_parties` | Perceived political conflict | 1 = Very strong … 4 = No conflict |
| `fav_eu` | Favourability of the European Union | 1 = Very favourable … 4 = Very unfavourable |
| `ideology` | Left–right self-placement | 0 = Extreme left … 6 = Extreme right |
| `d_class` | Subjective social class | 1 = Upper … 5 = Lower |

All variables except `ideology` follow the same convention: **lower numbers = more of the concept**. Reverse with `5 - variable` (for 4-point scales) or `4 - variable` (for 3-point scales) if you want higher = more.
:::

```{r}
#| label: ex6-your-question

# ── ✏️ Your group's analysis ──────────────────────────────────────────────────
# 1. Write your research question as a comment below.
# 2. Compute summary statistics using pew_svy and survey_mean().
# 3. Build a plot.

# Research question: [YOUR QUESTION HERE]

# ── Suggested scaffold ────────────────────────────────────────────────────────

# Step 1: Compute weighted summary statistics
# Replace YOUR_VARIABLE with the variable you want to analyse.
# If it is a 4-point scale: create a binary (e.g. YOUR_VARIABLE <= 2)
# or reverse it (e.g. 5 - YOUR_VARIABLE) in the mutate() below.

# my_summary <- pew_svy |>
#   filter(!is.na(YOUR_VARIABLE)) |>
#   group_by(country_f) |>
#   summarise(
#     pct_x = survey_mean(YOUR_VARIABLE <= 2, na.rm = TRUE, vartype = "se") * 100,
#     .groups = "drop"
#   )

# Step 2: Build a plot
# Choose the right geom for your question:
#   One variable across countries     → geom_col() + coord_flip()
#   Two variables, one dot per country → geom_point() + geom_text()
#   One variable split by two groups   → geom_col(position = "dodge") + facet_wrap()

# ggplot(my_summary, aes(x = reorder(country_f, pct_x), y = pct_x)) +
#   geom_col(fill = "#1B3A5C", width = 0.7) +
#   coord_flip() +
#   scale_y_continuous(labels = scales::label_percent(scale = 1)) +
#   labs(title = "YOUR TITLE", x = NULL, y = NULL,
#        caption = "Source: Pew Research Center Global Attitudes Spring 2024")
```

::: {.callout-caution}
## 💬 Your group's interpretation — write 2–3 sentences here

*State your research question, your key finding, and one limitation of your analysis.*

**Group answer:** *[write here]*
:::

---

# End-of-Session Reflection {.unnumbered}

::: {.callout-important}
## 🗣 Class discussion — pick one from your group

Each group shares **one finding** — the result that surprised you most, or the pattern you found hardest to explain. Keep it to 2–3 minutes per group.

Before presenting, agree on:
1. What did you find?
2. Why is it surprising (or not)?
3. What would you need to know to explain it better?
:::

::: {.callout-note collapse="true"}
## 📌 Navigator debrief — things to reflect on

Did the Driver understand each step, or did you explain and they typed without following? The test: can your Driver now write the `group_by() |> summarise(survey_mean(...))` pattern from memory?

Key concepts your Driver should be able to explain after today:

- [ ] Why we use `pew_svy` instead of `pew_clean` for estimates
- [ ] What `zap_missing()` does and why it matters before any analysis
- [ ] What the `_r` reversed variables mean and how they are created
- [ ] Why the mean of a binary (0/1) variable equals the proportion
- [ ] What `pivot_longer()` does and why it is needed before grouped plots
- [ ] Why weighting matters and what it changes in practice
:::

---

*Data: Pew Research Center Global Attitudes Spring 2024 Survey · Analysis: Survey Analytics and Big Data, Corvinus University of Budapest, 2025/2026*
