---
title: "Week_2_Intro_to_R"
format: html
editor: visual
---

## Creating a new project

## Project Setup in RStudio

Before you start analyzing data, it's important to organize your work in an **R Project**. This keeps all your files, scripts, and data together in one location.

### Step-by-Step Setup

1.  **Open RStudio** and go to **File** → **New Project**

2.  Select **New Directory** and choose **New Project**

3.  Give your project a meaningful name (e.g., `week_2_analysis` or `penguin_study`)

4.  Choose where you want to save it on your computer

5.  Click **Create Project** – RStudio will restart in your new project

### What Just Happened?

RStudio created a folder with: - A `.Rproj` file (the project file) - Your working directory is now set to this folder - All your scripts and data will live here

### Create Subfolders (Optional but recommended)

In the Files panel (bottom right), create these folders:

-   data/
-   scripts/
-   output/
-   misc/

## Prerequisites - installing packages

```{r}
#| eval: false
options(repos = c(CRAN = "https://cran.rstudio.com"))

install.packages("tidyverse")
install.packages("here") # this will make your life much easier when working with folders
install.packages("palmerpenguins")

```

```{r}
library(tidyverse)
library(here)
here()

```

## Let's have some data and play with it

### #1 Penguins

The Palmer Penguins dataset contains measurements and observations of three penguin species observed near Palmer Station in Antarctica. With 344 observations and realistic missing data.

```{r}
library(palmerpenguins) 
data(penguins)

head(penguins)
glimpse(penguins)
```

Save the data locally (so we no longer need to open the original)

```{r}
write_csv(penguins, here("data","penguins.csv"))
```

### Do a quick demo data visualization

Recreate the visualization displaying the relationship between flipper lengths and body masses of these penguins, taking into consideration the species of the penguin.

```{r}
ggplot(data = penguins)
```

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
)
```

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point()
```

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = "lm")
```

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = "lm")
```

```{r}
library(ggthemes)
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) +
  scale_color_colorblind()
```

A few other ideas (basic chart types) for visualization

```{r}
ggplot(penguins, aes(x = fct_infreq(species))) + # remove fct_infreq() and see what changed!
  geom_bar()
```

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200) # experiment with binwidth
```

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

```{r}
ggplot(penguins, aes(x = body_mass_g, color = species, fill = species)) +
  geom_density(alpha = 0.5)
```

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200) # experiment with binwidth
```

### How to save a plot?

```{r}
plot_1 <- ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) +
  scale_color_colorblind()

ggsave(filename = here("outputs","penguin-plot.png"), plot=plot_1) # saving in .png (bitmap)
ggsave(filename = here("outputs","penguin-plot.svg"), plot=plot_1) # saving in .svg (vector)

ggsave(filename = here("outputs","penguin-plot2.png"), plot=plot_1,
       width=7,
       height=4,
       dpi=600) # adjust the size


```

### 

### #2 Baby names (in the US)

Loading some interesting and funny sample data

We'll explore the dataset of US baby names. This dataset contains historical records of baby names given in the United States, spanning over a century. Let's learn how to load and examine this data.

First, we need to install and load the `babynames` package:

```{r}
#| eval: false
install.packages("babynames")
```

```{r}
# Load the package
library(babynames)

# Load the dataset into our environment
data(babynames)

head(babynames)
colnames(babynames)
glimpse(babynames)
```

Let's save this data locally similar to the previous one:

```{r}
write_csv(babynames, here("data","babynames.csv"))

# Create a bar chart of the TOP10 names in the babynames dataset
top10_names <- babynames %>%
  group_by(name) %>%
  summarise(total = sum(n)) %>%
  arrange(desc(total)) %>%
  slice_head(n = 10)

```
