---
title: "Data Practices 2026 — Assignment 02"
subtitle: "Country-Level Data Manipulation & Weighted Survey Analysis"
author: "Akbota Saduakassova" # change this to your own name 😉
date: April
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    code-fold: false
    df-print: paged
execute:
  echo: true
  eval: true
  warning: true
  message: true
editor: 
  markdown: 
    wrap: sentence
---

```{r}
#| label: setup-check
#| echo: false
#| eval: true
# This hidden chunk just confirms the document renders correctly.
# Students: ignore this block.
```

------------------------------------------------------------------------

# How to Use This Document {.unnumbered}

::: callout-important
## ✋ Read this first — five steps before you write a single line of code

**Step 1 — Open the right project.** This assignment assumes you are working inside an RStudio Project (see [Section 2](#sec-project) below).
Before anything else, make sure the correct `.Rproj` file is open.
You will see the project name in the top-right corner of RStudio.

**Step 2 — Run the Setup chunk.** Go to [Section 3](#sec-setup) and run the setup chunk.
It installs any missing packages, loads all libraries, and downloads both datasets.
Nothing in the exercises will work until this has run successfully.

**Step 3 — Find the ✏️ markers.** Every place where you need to write code or an answer is marked with a **✏️** icon.
Do not write code anywhere else unless you are experimenting in a scratch chunk.

**Step 4 — Use the hints.** Each exercise has two collapsible hint boxes.
Open **Hint 1** first — it gives you direction without giving away the solution.
Only open **Hint 2** if you are still stuck — it provides more detail and a worked example on a different dataset.

**Step 5 — Render before submitting.** When finished, click *Render* (Ctrl/Cmd + Shift + K).
Fix any errors until the document renders cleanly to HTML, then submit the `.html` file.
:::

------------------------------------------------------------------------

# Project Setup {#sec-project}

## Why use an RStudio Project?

When you open a plain R script or a `.qmd` file without a project, R has no idea where "home" is.
Every time you reference a file — a CSV, a plot output, a saved model — you either have to type out the full absolute path (`C:/Users/yourname/Documents/...`) or use `setwd()`, which breaks the moment someone else opens your code, or you move the folder.

An **RStudio Project** solves this by making the project folder the automatic working directory whenever you open it.
Your paths become relative to that folder, which means:

-   The same code works on your laptop, your colleague's laptop, and the grading server
-   You never need `setwd()` again
-   All your scripts, data, and outputs are self-contained in one place

::: {.callout-tip collapse="true"}
## 📁 Step-by-step: creating a new RStudio Project

If you already have a project from a previous week's class, you can use that — just skip to [Section 2.2](#sec-here).

**Creating a new project:**

1.  Open RStudio → `File` → `New Project`
2.  Choose **New Directory** → **New Project**
3.  Give it a meaningful name: `data_practices_2026` or `phd_assignment_02`
4.  Choose a location on your computer (e.g., `Documents/courses/`)
5.  Click **Create Project** — RStudio restarts inside the new folder

**Recommended folder structure (create these in the Files panel):**

```         
data_practices_2026/
├── data/          ← downloaded datasets live here
├── scripts/       ← standalone R scripts (optional)
├── output/        ← rendered HTML files, saved plots
└── assignment_02.qmd   ← this file
```

To create a subfolder: in the Files panel (bottom right), click **New Folder**.
:::

## The `here` package {#sec-here}

Once you are inside a project, the `here` package gives you a clean, OS-independent way to build file paths.
Instead of:

``` r
read_csv("C:/Users/myname/Documents/data_practices_2026/data/wdi.csv")  # fragile
read_csv("data/wdi.csv")   # works only if working directory is exactly right
```

You write:

``` r
read_csv(here("data", "wdi.csv"))   # always works, on any machine
```

`here()` automatically detects the project root (the folder with the `.Rproj` file) and builds the path from there.
It is especially important for Quarto documents, which sometimes set their own working directory differently from the console.

::: {.callout-note collapse="true"}
## 🔍 Why `here()` matters more than you think

Quarto documents render in their own environment.
When you click **Render**, R's working directory is set to the folder containing the `.qmd` file — which is *usually* the project root, but not always.
If you ever move the `.qmd` into a subfolder, all relative paths break immediately.
`here()` always anchors to the `.Rproj` file, so it survives reorganisation.

A practical rule: **never use `setwd()` inside a script or a `.qmd`. Use `here()` instead.**

You can verify it is working by running `here()` in the console — it should print your project folder path.
:::

------------------------------------------------------------------------

# Setup: Packages and Data {#sec-setup}

::: {.callout-note collapse="true"}
## 📦 What are packages and why do we need them? (Beginner explanation)

R comes with a set of built-in functions — enough to do basic arithmetic, read files, and fit simple models.
But the real power of R comes from **packages**: collections of functions written by researchers and developers around the world, freely shared via CRAN (the Comprehensive R Archive Network).

Think of packages like apps on a smartphone:

-   You **install** an app once (it downloads to your device)
-   You **open** it every time you want to use it

In R:

``` r
install.packages("tidyverse")   # install once — do this in the Console, not in a script
library(tidyverse)              # open every session — put this at the top of every script
```

**The packages we use in this assignment:**

| Package | What it does |
|------------------------------|------------------------------------------|
| `tidyverse` | A family of packages for data manipulation (`dplyr`), visualisation (`ggplot2`), and more |
| `WDI` | Downloads World Bank World Development Indicators data directly into R |
| `haven` | Reads SPSS `.sav` files (the Pew dataset format) |
| `srvyr` | Adds survey-weight support to tidyverse-style code |
| `here` | Builds file paths relative to the project root |
| `knitr` | Formats tables nicely when rendering |
| `scales` | Formats numbers and axes in plots (e.g. thousands separators, percentages) |

If you see an error like `there is no package called 'WDI'`, run the install line below in your **Console** (not in the document).
:::

```{r}
#| label: setup-packages
#| eval: true
#| message: false
#| warning: false

# Single source of truth for all required packages.
# Only installs what is not already present — safe to run repeatedly.
packages <- c("tidyverse", "WDI", "haven", "srvyr",
              "here", "knitr", "scales")

missing <- packages[!packages %in% installed.packages()[, "Package"]]
if (length(missing) > 0) install.packages(missing)

invisible(lapply(packages, library, character.only = TRUE))

```

## Download and prepare the WDI dataset

The **World Development Indicators (WDI)** is the World Bank's flagship database: over 1,500 indicators covering more than 200 countries from the 1960s to the present.
We pull it directly from the World Bank API using the `WDI` package — no manual download needed.

The data file is also available in the course GitHub repository as a backup: `https://github.com/sagvari/data_practices_2026/tree/main/R_course/data`

```{r}
#| label: download-wdi
#| eval: true
#| cache: true

# Pull a curated set of socioeconomic indicators for all countries, 2000-2022.
# extra = TRUE adds region, income group, and ISO3 code columns automatically.
wdi_raw <- WDI(
  indicator = c(
    "NY.GDP.PCAP.KD",   # GDP per capita (constant 2015 USD)
    "SP.DYN.LE00.IN",   # Life expectancy at birth (years)
    "SP.POP.TOTL",      # Total population
    "IT.NET.USER.ZS",   # Individuals using the internet (% of pop.)
    "SE.SEC.ENRR",      # Secondary school enrolment (% gross)
    "SL.UEM.TOTL.ZS",   # Unemployment rate (% of labour force)
    "SI.POV.GINI"       # Gini index (income inequality, 0–100)
  ),
  start = 2000,
  end   = 2022,
  extra = TRUE    # adds: region, income, iso3c, capital, longitude, latitude
)

# Rename indicator columns to human-friendly names, then clean up
wdi <- wdi_raw |>
  rename(
    country_name    = country,
    gdp_per_capita  = NY.GDP.PCAP.KD,
    life_expectancy = SP.DYN.LE00.IN,
    population      = SP.POP.TOTL,
    internet_use    = IT.NET.USER.ZS,
    school_enroll   = SE.SEC.ENRR,
    unemployment    = SL.UEM.TOTL.ZS,
    gini            = SI.POV.GINI
  ) |>
  filter(region != "Aggregates", !is.na(region)) |>
  select(country_name, iso2c, iso3c, year,
         region, income,
         gdp_per_capita, life_expectancy, population,
         internet_use, school_enroll,
         unemployment, gini)

# Save locally so you don't need to re-download on every render
write_csv(wdi, here("data", "wdi_indicators.csv"))

glimpse(wdi)
```

## Download and prepare the Pew 2024 dataset

The **Pew Research Center Global Attitudes Spring 2024** survey covers 24 countries with nationally representative samples and complex survey weights.

```{r}
#| label: download-pew
#| eval: true
#| cache: true

# Direct download from the course SharePoint repository
data_link <- "https://tkmtahu-my.sharepoint.com/:u:/g/personal/sagvari_bence_tk_hu/IQCc8k4_cII9TIlJmZQULNd6ATVy7R_EEkU5AwRBspfjP_A?download=1"

tmp <- tempfile(fileext = ".sav")
download.file(data_link, tmp, mode = "wb", quiet = TRUE)
pew_raw <- haven::read_spss(tmp)

# ── Recode and clean (matching Week 5 notebook) ───────────────────────────────
pew <- pew_raw |>
  mutate(
    # Country label (drop SPSS attributes)
    # country is numeric in the .sav; as_factor() uses the SPSS value labels
    country = as.character(as_factor(country)),

    # Survey weight
    weight  = as.numeric(weight),

    # Topic A — Democracy satisfaction (satisfied_democracy, Q3):
    #   1=Very satisfied, 2=Somewhat satisfied, 3=Not too satisfied, 4=Not at all satisfied
    dem_satisfied = case_when(
      satisfied_democracy %in% 1:2 ~ 1L,   # satisfied
      satisfied_democracy %in% 3:4 ~ 0L,   # not satisfied
      TRUE                         ~ NA_integer_
    ),

    # Topic A — Press freedom important (press_freedom, Q12a):
    #   1=Very important, 2=Somewhat important, 3=Not too important, 4=Not important at all
    press_freedom_imp = case_when(
      press_freedom %in% 1:2 ~ 1L,
      press_freedom %in% 3:4 ~ 0L,
      TRUE                   ~ NA_integer_
    ),

    # Topic B — Economic situation: national (econ_sit, Q1):
    #   1=Very good, 2=Somewhat good, 3=Somewhat bad, 4=Very bad
    econ_good = case_when(
      econ_sit %in% 1:2 ~ 1L,
      econ_sit %in% 3:4 ~ 0L,
      TRUE              ~ NA_integer_
    ),

    # Topic C — Religiosity: how important is religion (religion_import, Q23):
    #   1=Very important, 2=Somewhat important, 3=Not too important, 4=Not at all important
    religion_important = case_when(
      religion_import %in% 1:2 ~ 1L,
      religion_import %in% 3:4 ~ 0L,
      TRUE                     ~ NA_integer_
    ),

    # Demographic: age group (age, Q109a — numeric age in years)
    age_group = case_when(
      age < 30          ~ "18–29",
      age < 50          ~ "30–49",
      age < 65          ~ "50–64",
      age >= 65         ~ "65+",
      TRUE              ~ NA_character_
    ),

    # Demographic: gender is already named `gender` in the dataset (Q108)
    # Values: 1 = Male, 2 = Female — recode to character labels
    gender_label = case_when(
      gender == 1 ~ "Male",
      gender == 2 ~ "Female",
      TRUE        ~ NA_character_
    )
  ) |>
  filter(!is.na(weight)) |>
  select(country, weight, dem_satisfied, press_freedom_imp,
         econ_good, religion_important, age_group,
         gender = gender_label)   # rename for cleaner downstream use

# Save locally
write_csv(pew, here("data", "pew2024_cleaned.csv"))

glimpse(pew)
```

```{r}
#| label: survey-design
#| eval: true

# Create the srvyr survey design object — required for all weighted analyses
pew_design <- pew |>
  as_survey_design(weights = weight)
```

------------------------------------------------------------------------

# Part 1 — Country-Level Data Manipulation with WDI {#sec-wdi}

In this section you will use the WDI dataset to practise the core `dplyr` verbs covered in Weeks 2 and 3.
The dataset has **`r if(exists("wdi")) nrow(wdi) else "~30,000"`** rows (country × year observations) and 14 columns.

::: callout-note
## 🗂️ Quick variable reference

| Variable | Description | Unit |
|------------------------|-------------------------------|-----------------|
| `country_name` | Country name | — |
| `iso3c` | ISO 3-letter country code | — |
| `year` | Year | — |
| `region` | World Bank region | — |
| `income` | Income group (Low / Lower-middle / Upper-middle / High) | — |
| `gdp_per_capita` | GDP per capita | Constant 2015 USD |
| `life_expectancy` | Life expectancy at birth | Years |
| `population` | Total population | Persons |
| `internet_use` | Internet users | \% of population |
| `school_enroll` | Secondary school enrolment | \% gross |
| `unemployment` | Unemployment rate | \% of labour force |
| `gini` | Gini coefficient (income inequality) | 0 = perfect equality, 100 = maximum inequality |
:::

------------------------------------------------------------------------

## ✏️ Exercise 1 — Filtering: Who Broke 80 Years? {#sec-ex1}

**Topic:** Life expectancy and the high-income world

The global average life expectancy was around 73 years in 2022.
A life expectancy above 80 is often taken as a marker of a well-functioning health system and high living standards.

**Your task:** Using the most recent year in the dataset (2022), find all **high-income** countries where life expectancy exceeded **80 years**.
Sort the result from highest to lowest life expectancy and display the country name, region, life expectancy, and GDP per capita.

```{r}
#| label: ex1-solution
#| dependson: "download-wdi"

result <- wdi |>
  filter(
    year == 2022,
    income == "High income",
    life_expectancy > 80
  ) |>
  arrange(desc(life_expectancy)) |>
  mutate(
    life_expectancy = round(life_expectancy, 1),
    gdp_per_capita = round(gdp_per_capita, 0)
  ) |>
  select(
    Country = country_name,
    Region = region,
    `Life expectancy` = life_expectancy,
    `GDP per capita` = gdp_per_capita
  )

knitr::kable(result)
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

You need to narrow down the rows in two ways: by year and by a numeric threshold.
Think about which `dplyr` verb keeps only the rows that match certain conditions.
You will need to apply multiple conditions at once — remember that a comma inside `filter()` acts like AND.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

Use `filter()` with multiple conditions, then `arrange()` to sort:

``` r
# Example: countries with population over 100 million in 2022
wdi |>
  filter(year == 2022, population > 100e6) |>
  arrange(desc(population))
```

For your task, replace `population > 100e6` with the right variable and threshold.
Don't forget to also filter by income group — check the exact label by running `unique(wdi$income)` first.
:::

------------------------------------------------------------------------

## ✏️ Exercise 2 — Selecting and Renaming: A Clean Country Snapshot {#sec-ex2}

**Topic:** Internet access in Central and Eastern Europe

You want to share a clean, readable table of internet access figures with a non-technical colleague who does not need to see all 14 columns.

**Your task:** From the 2019 data (pre-pandemic, the last "normal" year for many indicators), select only countries in the **Europe & Central Asia** region.
Keep only four columns: country name, year, internet use, and unemployment.
Rename them to something human-friendly (`Country`, `Year`, `Internet use (%)`, `Unemployment (%)`).
Sort by internet use, descending.

```{r}
#| label: ex2-solution
#| dependson: "download-wdi"

wdi |>
  filter(
    year == 2019,
    region == "Europe & Central Asia"
  ) |>
  select(
    Country = country_name,
    Year = year,
    `Internet use (%)` = internet_use,
    `Unemployment (%)` = unemployment
  ) |>
  arrange(desc(`Internet use (%)`)) |>
  knitr::kable()
# ✏️ Write your solution here

```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

This is a filtering + column selection task.
Think: which verb subsets **rows**?
Which verb subsets (and renames) **columns**?
You can rename inside `select()` using the `new_name = old_name` syntax.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

`select()` lets you rename on the fly:

``` r
# Example: rename columns while selecting them
babynames |>
  select(Year = year, Name = name, Count = n)
```

Chain `filter()` before `select()` to first narrow down the rows.
Check the exact region label with `unique(wdi$region)`.
:::

------------------------------------------------------------------------

## ✏️ Exercise 3 — Mutate: The Digital Divide Index {#sec-ex3}

**Topic:** Computing a derived indicator from two existing variables

Economists and sociologists often talk about the **digital divide**: the gap between rich and poor countries in access to the internet.
One simple way to quantify how *efficiently* a country converts economic wealth into internet access is to compare internet use against GDP per capita.
A country with high internet use *relative to* its GDP is punching above its weight; a country with low internet use *relative to* its GDP has room to catch up.

**Your task:** Using the 2019 data (a stable pre-pandemic year):

1.  Create a new column called `internet_per_gdp` defined as `internet_use` divided by `(gdp_per_capita / 1000)` — this rescales GDP so the resulting ratio is on a readable scale.
2.  Remove rows where either variable is missing.
3.  Show the **10 countries with the highest ratio** (high internet access relative to GDP) and the **10 countries with the lowest ratio** (low internet access relative to GDP) in a single combined table. Add a column `group` labelling each country as `"High"` or `"Low"` so they are easy to distinguish.
4.  What does the pattern in the "Low" group tell you about the relationship between economic development and internet access?
High rates of Internet access are not always associated with higher GDP because Internet access requires certain infrastructure.
```{r}
#| label: ex3-solution
#| dependson: "download-wdi"


# ✏️ Write your solution here
data_2019 <- wdi |>
  filter(year == 2019) |>
  mutate(internet_per_gdp = internet_use / (gdp_per_capita / 1000)) |>
  filter(!is.na(internet_use), !is.na(gdp_per_capita), !is.na(internet_per_gdp))

top10 <- data_2019 |>
  slice_max(internet_per_gdp, n = 10) |>
  mutate(group = "High")

bottom10 <- data_2019 |>
  slice_min(internet_per_gdp, n = 10) |>
  mutate(group = "Low")

result <- bind_rows(top10, bottom10) |>
  mutate(
    internet_use = round(internet_use, 1),
    gdp_per_capita = round(gdp_per_capita, 0),
    internet_per_gdp = round(internet_per_gdp, 2)
  ) |>
  select(
    Country = country_name,
    Region = region,
    `Internet use (%)` = internet_use,
    `GDP per capita` = gdp_per_capita,
    `Internet per GDP` = internet_per_gdp,
    Group = group
  )

knitr::kable(result)
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

There is a `dplyr` verb that adds a new column to a data frame without removing any existing ones — it works like a calculator that can see all the other columns.
After creating the new column, think about how to remove rows where the result could not be computed (i.e. where one of the inputs was `NA`).
To show the top and bottom rows together, you will need to combine two separate results.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

`mutate()` creates the new column.
`filter(!is.na(variable))` removes rows where it is missing.
`slice_max()` and `slice_min()` extract the top/bottom *n* rows.
`bind_rows()` stacks two data frames vertically:

``` r
# Example: create a ratio, then show top-5 and bottom-5 combined
data_2019 <- wdi |>
  filter(year == 2019) |>
  mutate(school_per_gdp = school_enroll / (gdp_per_capita / 1000)) |>
  filter(!is.na(school_per_gdp))

top5    <- slice_max(data_2019, school_per_gdp, n = 5) |> mutate(group = "High")
bottom5 <- slice_min(data_2019, school_per_gdp, n = 5) |> mutate(group = "Low")

bind_rows(top5, bottom5) |>
  select(country_name, region, school_per_gdp, group)
```

Apply the same logic to `internet_per_gdp` and adjust *n* to 10.
:::

------------------------------------------------------------------------

## ✏️ Exercise 4 — Group By and Summarise: Regional Inequality Profiles {#sec-ex4}

**Topic:** Income inequality across world regions

The Gini coefficient measures income inequality: 0 means everyone earns the same, 100 means one person earns everything.
The WDI includes Gini estimates for most countries, though with many missing values because inequality surveys are conducted infrequently.

**Your task:** Using the most recent available Gini value **per country** (some countries have data in 2021, others only in 2018 — use the latest non-missing value per country), compute the **median and mean Gini** by World Bank `region`, along with the **number of countries** included in each regional average.
Sort by median Gini, descending.

*Guidance: This is a multi-step problem. First get the latest Gini per country, then summarise by region.*

```{r}
#| label: ex4-solution
#| dependson: "download-wdi"

# ✏️ Write your solution here
data_2019 <- wdi |>
  filter(year == 2019) |>
  mutate(internet_per_gdp = internet_use / (gdp_per_capita / 1000)) |>
  filter(!is.na(internet_use), !is.na(gdp_per_capita), !is.na(internet_per_gdp))

top10 <- data_2019 |>
  slice_max(internet_per_gdp, n = 10) |>
  mutate(group = "High")

bottom10 <- data_2019 |>
  slice_min(internet_per_gdp, n = 10) |>
  mutate(group = "Low")

bind_rows(top10, bottom10) |>
  select(country_name, region, internet_use, gdp_per_capita, internet_per_gdp, group) |>
  knitr::kable()
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

This requires two separate grouping operations.
First: group by country to find the most recent non-missing Gini year per country.
Second: group by region to compute summary statistics.
Think about `filter(!is.na(...))`, `group_by()`, `slice_max()`, and `summarise()`.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

To get the latest non-missing value per country, you can filter out NAs, group by country, and then take the row with the maximum year:

``` r
# Example: latest non-missing unemployment per country
wdi |>
  filter(!is.na(unemployment)) |>
  group_by(country_name) |>
  slice_max(year, n = 1) |>
  ungroup()
```

Then pipe that result into a second `group_by(region) |> summarise(...)` step.
Inside `summarise()`, use `median(..., na.rm = TRUE)`, `mean(..., na.rm = TRUE)`, and `n()`.
:::

------------------------------------------------------------------------

## ✏️ Exercise 5 — Multi-step Pipeline with Visualisation: The Health-Wealth Relationship {#sec-ex5}

**Topic:** Does money buy health?
The Preston Curve

The **Preston Curve** is one of the most replicated findings in demography: richer countries tend to have higher life expectancy, but the relationship is concave (diminishing returns).
Your task is to reproduce a version of this classic chart.

**Your task:**

1.  Take the 2019 data (before pandemic disruptions).
2.  Keep only countries with non-missing values for both `gdp_per_capita` and `life_expectancy`.
3.  Remove countries with population under 1 million (small island states distort the picture).
4.  Create a scatter plot with **log-scaled GDP per capita** on the x-axis and **life expectancy** on the y-axis. Colour the points by `income` group. Add a smooth trend line (not broken by group). Label at minimum the five richest and five poorest countries by name using `geom_text()` or `ggrepel`.
5.  Add informative axis labels, a title, and a caption citing the data source.

```{r}
#| label: ex5-solution
#| fig-width: 10
#| fig-height: 6
#| dependson: "download-wdi"

# ✏️ Write your solution here

data_2019 <- wdi |>
  filter(
    year == 2019,
    !is.na(gdp_per_capita),
    !is.na(life_expectancy),
    population >= 1e6
  )

label_data <- bind_rows(
  data_2019 |> slice_max(gdp_per_capita, n = 5),
  data_2019 |> slice_min(gdp_per_capita, n = 5)
)

ggplot(data_2019, aes(x = gdp_per_capita, y = life_expectancy, colour = income)) +
  geom_point(alpha = 0.7) +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE, colour = "black") +
  geom_text(
    data = label_data,
    aes(label = country_name),
    size = 3,
    vjust = -0.5,
    show.legend = FALSE
  ) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    title = "The Health-Wealth Relationship, 2019",
    subtitle = "Life expectancy rises with GDP per capita, but with diminishing returns",
    x = "GDP per capita (log scale, constant 2015 USD)",
    y = "Life expectancy at birth (years)",
    colour = "Income group",
    caption = "Source: World Bank, World Development Indicators (WDI)"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

This is a full pipeline: filter → mutate (optional log transform) → plot.
For the scatter plot, `ggplot2` handles this naturally with `geom_point()`.
The log scale can be applied with `scale_x_log10()` rather than transforming the variable itself.
For country labels, look at the `ggrepel` package (`geom_text_repel()`), which avoids overlapping labels — install it if needed.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

Basic scatter plot structure with a log x-axis and a smooth line:

``` r
# Example structure
ggplot(data, aes(x = some_variable, y = other_variable, colour = group)) +
  geom_point(alpha = 0.7) +
  geom_smooth(aes(group = 1), method = "loess", colour = "grey30", se = FALSE) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_colour_viridis_d(option = "plasma") +
  labs(title = "...", x = "...", y = "...", caption = "Source: World Bank WDI")
```

To label only extreme countries, create a separate `label_data` object filtered to `slice_max(gdp_per_capita, n=5)` and `slice_min(gdp_per_capita, n=5)`, then add `geom_text(data = label_data, aes(label = country_name))`.
:::

------------------------------------------------------------------------

# Part 2 — Weighted Survey Analysis with Pew 2024 {#sec-pew}

In this section you will work with the **Pew Research Center Global Attitudes Spring 2024** survey.
All estimates must use the survey weight variable to be nationally representative.
We use the `srvyr` package, which brings familiar tidyverse syntax to weighted survey analysis.

::: callout-important
## ⚖️ Why weights are not optional

The Pew sample is not a simple random sample.
Some groups (e.g. younger adults, minority language speakers) are deliberately over- or under-sampled to ensure enough cases for subgroup analysis, and to compensate for differential non-response rates.
The `weight` variable corrects for this so that your estimates represent the **national adult population**, not the people who happened to respond.

Running `mean(pew$dem_satisfied, na.rm = TRUE)` would give you the *sample* proportion — which can be several percentage points away from the nationally representative estimate.
Always use `pew_design` with `srvyr` functions.

``` r
# Correct — weighted
pew_design |> summarise(survey_mean(dem_satisfied, na.rm = TRUE))

# Wrong for inference — unweighted
pew |> summarise(mean(dem_satisfied, na.rm = TRUE))
```
:::

------------------------------------------------------------------------

## ✏️ Exercise 6 — Weighted Frequencies: Democracy Satisfaction by Country {#sec-ex6}

**Topic:** How satisfied are people with democracy?

**Your task:** For each country in the dataset, calculate the **weighted percentage of respondents who are satisfied with democracy** (`dem_satisfied == 1`).
Sort the results from highest to lowest satisfaction.
Display as a clean table showing country, percentage satisfied, and the effective (unweighted) number of respondents per country.

```{r}
#| label: ex6-solution
#| dependson: "download-pew"


# ✏️ Write your solution here
pew_design |>
  group_by(country) |>
  summarise(
    pct_satisfied = survey_mean(dem_satisfied, na.rm = TRUE) * 100,
    n = unweighted(n())
  ) |>
  arrange(desc(pct_satisfied)) |>
  select(
    Country = country,
    `SAtisfied Rate` = pct_satisfied,
    `Respondents (n)` = n
  ) |>
  knitr::kable(digits = 1)
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

You need to group the survey design object by country, then compute a weighted mean of a binary 0/1 variable.
The weighted mean of a binary variable equals the weighted proportion of 1s.
All `srvyr` summarise functions work inside `summarise()` just like regular `dplyr` — but use `survey_mean()` instead of `mean()`.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

``` r
# Example: weighted % saying religion is important, by country
pew_design |>
  group_by(country) |>
  summarise(
    pct_religious = survey_mean(religion_important, na.rm = TRUE),
    n = unweighted(n())
  ) |>
  arrange(desc(pct_religious))
```

`survey_mean()` returns the proportion by default.
Multiply by 100 for a percentage.
`unweighted(n())` gives the actual number of survey respondents (not the weighted count).
:::

------------------------------------------------------------------------

## ✏️ Exercise 7 — Weighted Crosstab: Press Freedom Importance by Age Group {#sec-ex7}

**Topic:** Generational differences in attitudes toward press freedom

**Your task:** Across all countries pooled, compute the **weighted percentage who consider press freedom important** (`press_freedom_imp == 1`) **separately for each age group** (`age_group`).
Then break this down further by `gender`.
Present the result as a crosstab with age groups in rows and gender in columns.
Write two sentences interpreting the pattern.
Seems like women consider freedom of press more important than men. Across all countries age group of 65+ consider freedom of press comparatively less important than younger age groups.
```{r}
#| label: ex7-solution
#| dependson: "download-pew"

# ✏️ Write your solution here
pew_design |>
  filter(!is.na(country), !is.na(age_group), !is.na(gender)) |>
  group_by(country, age_group, gender) |>
  summarise(
    pct_press_freedom = survey_mean(press_freedom_imp, na.rm = TRUE) * 100
  ) |>
  mutate(pct_press_freedom = round(pct_press_freedom, 1)) |>
  select(country, age_group, gender, pct_press_freedom) |>
  pivot_wider(
    names_from = gender,
    values_from = pct_press_freedom
  ) |>
  arrange(country, age_group) |>
  knitr::kable(
    col.names = c("Country", "Age group", "Female", "Male")
  )
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

This is a two-way breakdown: group by two variables simultaneously and compute a weighted proportion.
The result will be in long format — think about how to pivot it into a wide crosstab with one column per gender.
Look at `pivot_wider()` for reshaping.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

``` r
# Example: weighted proportions by two grouping variables, then pivot wide
pew_design |>
  filter(!is.na(age_group), !is.na(gender)) |>
  group_by(age_group, gender) |>
  summarise(pct = survey_mean(econ_good, na.rm = TRUE)) |>
  select(age_group, gender, pct) |>
  pivot_wider(names_from = gender, values_from = pct)
```

Multiply `pct` by 100 and round to 1 decimal before pivoting for a cleaner table.
You can pass the result to `knitr::kable()` for nice formatting.
:::

------------------------------------------------------------------------

## ✏️ Exercise 8 — Comparing Weighted Means: Economic Optimism Across Income Levels {#sec-ex8}

**Topic:** Does economic context shape personal economic sentiment?

Pew surveys multiple countries at very different income levels.
This exercise asks whether people in richer countries feel better about their personal economic situation than people in poorer countries — after accounting for survey weights.

**Your task:**

1.  Join the Pew data with the WDI 2023 GDP per capita figures (match on country name — you will need to handle some name mismatches manually or use `stringr` to clean them).
2.  Create a variable `gdp_group` that splits countries into three roughly equal groups by GDP per capita: Low, Middle, High (use `ntile()` or manual cut-points).
3.  Compute the **weighted percentage reporting a good personal economic situation** (`econ_good == 1`) by `gdp_group`.
4.  Visualise the result as a simple bar chart with error bars (use the standard error returned by `survey_mean()`).

```{r}
#| eval: true
#| label: ex8-solution
#| fig-width: 7
#| fig-height: 4
#| dependson: ["download-wdi", "download-pew"]

# ✏️ Write your solution here


#1
wdi_2023 <- WDI(
  indicator = "NY.GDP.PCAP.KD",
  start = 2023,
  end = 2023,
  extra = TRUE
) |>
  rename(
    country_name = country,
    gdp_per_capita = NY.GDP.PCAP.KD
  ) |>
  filter(region != "Aggregates", !is.na(region), !is.na(gdp_per_capita)) |>
  select(country_name, gdp_per_capita)

#2
pew_joined <- pew |>
  mutate(
    country_join = recode(
      country,
      "US" = "United States",
      "South Korea" = "Korea, Rep.",
      "Ivory Coast" = "Cote d'Ivoire",
      "Czech Republic" = "Czechia",
      "Russia" = "Russian Federation",
      "Turkey" = "Turkiye",
      .default = country
    )
  ) |>
  left_join(wdi_2023, by = c("country_join" = "country_name"))


pew_joined |>
  filter(is.na(gdp_per_capita)) |>
  distinct(country)
# 3
country_gdp_groups <- pew_joined |>
  distinct(country, gdp_per_capita) |>
  filter(!is.na(gdp_per_capita)) |>
  mutate(
    gdp_group = ntile(gdp_per_capita, 3),
    gdp_group = case_when(
      gdp_group == 1 ~ "Low",
      gdp_group == 2 ~ "Middle",
      gdp_group == 3 ~ "High"
    )
  )

# 4
pew_with_gdp <- pew_joined |>
  left_join(country_gdp_groups, by = c("country", "gdp_per_capita"))

# 5
pew_design_gdp <- pew_with_gdp |>
  as_survey_design(weights = weight)

# 6
econ_by_gdp <- pew_design_gdp |>
  filter(!is.na(gdp_group)) |>
  group_by(gdp_group) |>
  summarise(
    result = survey_mean(econ_good, na.rm = TRUE, vartype = "se")
  ) |>
  mutate(
    pct = result * 100,
    se_pct = result_se * 100,
    lower = pct - 1.96 * se_pct,
    upper = pct + 1.96 * se_pct
  )

# 7. Table
econ_by_gdp |>
  select(
    `GDP group` = gdp_group,
    `Percent saying economy is good` = pct,
    `SE` = se_pct
  ) |>
  knitr::kable(digits = 1)
# 8. Bar chart +
ggplot(econ_by_gdp, aes(x = gdp_group, y = pct)) +
  geom_col() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  labs(
    title = "Personal economic sentiment by country GDP group",
    subtitle = "Weighted percentage reporting a good personal economic situation",
    x = "GDP per capita group",
    y = "Rate saying economy is good",
    caption = "Pew Global Attitudes Spring 2024; World Bank WDI 2023 GDP per capita"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

This requires a join between two datasets.
The challenge is that country names in Pew may not exactly match those in WDI (e.g. "United States" vs "United States of America").
Think about which join function to use, and how to check which countries failed to match.
For the visualisation, `geom_col()` with `geom_errorbar()` is the standard approach for bar charts with confidence intervals.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

``` r
# Get WDI country names for a specific year
wdi_2022 <- wdi |>
  filter(year == 2022, !is.na(gdp_per_capita)) |>
  select(country_name, gdp_per_capita)

# Left join — rows in pew without a match get NA
pew_joined <- pew |>
  left_join(wdi_2022, by = c("country" = "country_name"))

# Check which countries didn't match
pew_joined |> filter(is.na(gdp_per_capita)) |> distinct(country)
```

For the error bars, `survey_mean()` returns both `_mean` and `_se` columns when you name the output:

``` r
pew_design |>
  group_by(gdp_group) |>
  summarise(result = survey_mean(econ_good, na.rm = TRUE, vartype = "se"))
# This gives result (the mean) and result_se (the standard error)
```

Use `geom_errorbar(aes(ymin = result - 1.96*result_se, ymax = result + 1.96*result_se))`.
:::

------------------------------------------------------------------------

## ✏️ Exercise 9 — Weighted Subgroup Analysis: Religion and Economic Views {#sec-ex9}

**Topic:** Does religiosity correlate with economic optimism?

**Your task:** Among European countries only (select at least 5 countries from the Pew data that are clearly European), test whether **religiously important respondents** differ from **non-religious respondents** in their assessment of the **personal economic situation** (`econ_good`).
Compute the weighted proportions for each religious group separately **per country**, then visualise the comparison as a **grouped bar chart** or a **dot plot with connecting lines** (one dot per country, connected across religious groups).
Write a short interpretation.

```{r}
#| eval: true
#| label: ex9-solution
#| fig-width: 9
#| fig-height: 5
#| dependson: "ex7-solution"

# ✏️ Write your solution here

result <- pew_design |>
  filter(
    country %in% c("Germany", "Slovakia", "Monaco", "Sweden", "Spain", "Hungary"),
    !is.na(religion_important),
    !is.na(econ_good)
  ) |>
  group_by(country, religion_important) |>
  summarise(
    pct = survey_mean(econ_good, na.rm = TRUE)
  ) |>
  mutate(
    pct = pct * 100,
    religious_label = ifelse(
      religion_important == 1,
      "Religion important",
      "Religion not important"
    )
  )

result |>
  knitr::kable(digits = 1)

ggplot(
  result,
  aes(
    x = pct,
    y = reorder(country, pct),
    colour = religious_label
  )
) +
  geom_line(aes(group = country), colour = "grey70") +
  geom_point(size = 3) +
  labs(
    title = "Religiosity and personal economic optimism in selected European countries",
    subtitle = "Weighted percentage reporting a good personal economic situation",
    x = "Share reporting personal economic situation is good",
    y = "Country",
    colour = NULL,
    caption = "Source: Pew Global Attitudes Spring 2024"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

Filter `pew_design` to European countries first using `filter(country %in% c(...))`.
Then group by two variables: country and religiosity.
Compute the weighted proportion of `econ_good` for each combination.
A dot plot with lines connecting the two religiosity dots within each country is effective — think about `geom_point()` and `geom_line()`, with country on one axis and proportion on the other.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

``` r
# Compute weighted proportions by two grouping variables
result <- pew_design |>
  filter(country %in% c("Germany", "France", "Poland",
                         "Sweden", "Spain", "Hungary"),
         !is.na(religion_important), !is.na(econ_good)) |>
  group_by(country, religion_important) |>
  summarise(pct = survey_mean(econ_good, na.rm = TRUE)) |>
  mutate(religious_label = ifelse(religion_important == 1,
                                  "Religion important", "Not important"))
```

For the connected dot plot:

``` r
ggplot(result, aes(x = pct, y = reorder(country, pct),
                   colour = religious_label)) +
  geom_line(aes(group = country), colour = "grey70") +
  geom_point(size = 4) +
  scale_x_continuous(labels = scales::percent) +
  labs(colour = NULL)
```
:::

------------------------------------------------------------------------

## ✏️ Exercise 10 — Advanced: A Country Ranking Dashboard {#sec-ex10}

**Topic:** Building a multi-indicator summary

**Your task:** Produce a single summary table that ranks all Pew countries by **four weighted indicators simultaneously**:

1.  `dem_satisfied` — % satisfied with democracy
2.  `press_freedom_imp` — % valuing press freedom
3.  `econ_good` — % with positive economic outlook
4.  `religion_important` — % for whom religion is important

For each indicator, compute the weighted national proportion.
Then add a `composite_score` column defined as the **mean of the first three indicators** (democracy satisfaction + press freedom + economic optimism) — treating this as a rough "liberal-democratic optimism index".
Sort by `composite_score` descending.

Finally, visualise the result as a **heatmap** where rows are countries, columns are the four indicators, and the fill colour encodes the weighted percentage.
Use a diverging colour scale centred at 50%.

```{r}
#| eval: true
#| label: ex10-solution
#| fig-width: 10
#| fig-height: 7
#| dependson: "ex7-solution"

# ✏️ Write your solution here

summary_table <- pew_design |>
  group_by(country) |>
  summarise(
    dem_sat   = survey_mean(dem_satisfied, na.rm = TRUE),
    press_fr  = survey_mean(press_freedom_imp, na.rm = TRUE),
    econ_opt  = survey_mean(econ_good, na.rm = TRUE),
    religious = survey_mean(religion_important, na.rm = TRUE)
  ) |>
  mutate(
    composite_score = (dem_sat + press_fr + econ_opt) / 3
  ) |>
  mutate(
    dem_sat = round(dem_sat * 100, 1),
    press_fr = round(press_fr * 100, 1),
    econ_opt = round(econ_opt * 100, 1),
    religious = round(religious * 100, 1),
    composite_score = round(composite_score * 100, 1)
  ) |>
  arrange(desc(composite_score))

summary_table |>
  select(
    Country = country,
    `Democracy satisfied (%)` = dem_sat,
    `Press freedom important (%)` = press_fr,
    `Economic optimism (%)` = econ_opt,
    `Religion important (%)` = religious,
    `Composite score (%)` = composite_score
  ) |>
  knitr::kable()

summary_long <- summary_table |>
  select(country, dem_sat, press_fr, econ_opt, religious) |>
  pivot_longer(
    cols = -country,
    names_to = "indicator",
    values_to = "pct"
  ) |>
  mutate(
    indicator = recode(
      indicator,
      dem_sat = "Democracy satisfied",
      press_fr = "Press freedom",
      econ_opt = "Economic optimism",
      religious = "Religion important"
    ),
    country = factor(country, levels = summary_table$country)
  )

ggplot(summary_long, aes(x = indicator, y = country, fill = pct)) +
  geom_tile() +
  geom_text(aes(label = paste0(round(pct, 0), "%")), size = 3) +
  scale_fill_gradient2(
    low = "#c84b2f",
    mid = "#f5f5f5",
    high = "#2a6b6e",
    midpoint = 50
  ) +
  labs(
    title = "Weighted national estimates — Pew 2024 Global Attitudes",
    subtitle = "Countries ranked by composite liberal-democratic optimism score",
    x = NULL,
    y = NULL,
    fill = "%"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

::: {.callout-tip collapse="true"}
## 💡 Hint 1 — Where to start

This requires four separate `summarise()` calls (one per indicator) or a clever `across()` trick.
The cleanest approach is to compute each indicator in one `group_by(country) |> summarise(...)` call.
Then reshape the result to long format for the heatmap using `pivot_longer()`.
For `geom_tile()`, the fill aesthetic should be the weighted proportion, with a diverging palette.
:::

::: {.callout-tip collapse="true"}
## 💡 Hint 2 — More detail

Computing multiple indicators in one `summarise()` call:

``` r
summary_table <- pew_design |>
  group_by(country) |>
  summarise(
    dem_sat   = survey_mean(dem_satisfied,     na.rm = TRUE),
    press_fr  = survey_mean(press_freedom_imp, na.rm = TRUE),
    econ_opt  = survey_mean(econ_good,         na.rm = TRUE),
    religious = survey_mean(religion_important, na.rm = TRUE)
  ) |>
  # keep only the point estimates, drop SE columns
  select(country, dem_sat, press_fr, econ_opt, religious)
```

For the heatmap, pivot to long format first:

``` r
summary_long <- summary_table |>
  pivot_longer(-country, names_to = "indicator", values_to = "pct")

ggplot(summary_long, aes(x = indicator, y = reorder(country, pct),
                          fill = pct * 100)) +
  geom_tile() +
  scale_fill_gradient2(low = "#c84b2f", mid = "#f5f5f5",
                        high = "#2a6b6e", midpoint = 50) +
  geom_text(aes(label = paste0(round(pct*100, 0), "%")),
            size = 3, colour = "grey20") +
  labs(fill = "%", x = NULL, y = NULL,
       title = "Weighted national estimates — Pew 2024 Global Attitudes") +
  theme_minimal()
```
:::

------------------------------------------------------------------------

# Submission Checklist {.unnumbered}

Before submitting, check each item:

-   [ ] All 10 exercises have code in the ✏️ chunks
-   [ ] The document renders to HTML without errors (`Render` button or Ctrl+Shift+K)
-   [ ] Exercises 5, 8, 9, and 10 include a plot
-   [ ] Exercises 7, 9, and 10 include a written interpretation (2–3 sentences)
-   [ ] You are submitting the rendered **`.html`** file, not the `.qmd`
-   [ ] Your name and student ID appear on the rendered document (add them to the `author:` field in the YAML header above)

------------------------------------------------------------------------

*Data sources: World Bank World Development Indicators (CC-BY 4.0) · Pew Research Center Global Attitudes Spring 2024 Survey · Corvinus University of Budapest, Survey Analytics and Big Data, 2025/2026*
